---
title: "R packages"
---

```{r setup3r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = NA, fig.path = "figure")
```


In the previous sections we have used 'built in' R functions such as `mean`, `max` and `min`. 

R packages contain additional functions, written and developed by the R community. These packages can contain data sets and functions to perform specialised analysis or graphics. 

To download an R package, click on Packages in the Files / Plots / Packages / Help panel, then click on Install package(s). The default option will be to download packages from ‘Repository (CRAN)’, this is where many R packages are uploaded to. In the ‘Packages’ box you can type the name of the package you want to download. 

Try downloading the package `{RColorBrewer}` by typing the package name into the ‘Packages’ box and clicking ‘Install’. 

After downloading an R package it will be listed under Packages in the Files / Plots / Packages / Help panel. To use the functions in a package, we must load the R package using `library(packagename)`.

Type `library(RColorBrewer)` into your R script and run the line. The package will now be loaded and you can use the functions the package contains. .

::: {.callout-note icon="false}

## Exercise

A colleague has sent you some code to make a map of Brazil, but when you run their code you see that you don't have the required package installed. 
```{r package, echo=TRUE, error = TRUE}
map("world", "Brazil") # Map of Brazil
map.cities(country = "Brazil") # Add a point for each city
```

Using a search engine, find the package, install it, and run the code.
:::

::: {.callout-note collapse="true" icon="false}
## Solution

```{r package2, echo = TRUE, warning = FALSE}
library("maps")
map("world", "Brazil")
map.cities(country = "Brazil")
```

:::
