---
title: "Basic plots"
---

```{r setup3r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = NA, fig.path = "figure")
```

R can be used for data analysis and manipulating data. In both of these cases, we want to be able to plot our data.

The code below loads in the data set `women` and makes a basic scatter plot of height against weight. Copy and paste this R code into your R script then run the lines. The plot below will appear in your Files / Plots / Packages / Help  panel. 

```{r plot1}
data("women")
plot(women$weight, women$height)
```

At the moment, our plot isn't easy to read, the axis labels are not clear. We can improve it by changing our axis labels using `xlab` for changing the x axis label and `ylab` for changing the y axis label. 

```{r plot2, echo=TRUE}
plot(women$weight, women$height,
     xlab = "Weight (lbs)", ylab = "Height (in)")
```

Let's also add a main title, to do this we use the argument `main`.
```{r plot3, echo=TRUE}
plot(women$weight, women$height,
     xlab = "Weight (lbs)", ylab = "Height (in)",
     main = "Plot of weight against height")
```

We can change the colour of the points using `col = "blue"` and change the shape of the points using `pch`. 
```{r plot4, echo=TRUE}
plot(women$weight, women$height,
     xlab = "Weight (lbs)", ylab = "Height (in)",
     main = "Plot of weight against height",
     pch = 19, col = "blue")
```


To save our plot, we can click on Export (just above the plot) and either save as Image or PDF. Select as Image. You change the name of the file in the 'File Name:' box.

If you selected a folder to be your working directory, then the plot should appear in that folder. To check what your current working director is type `getwd()` into the console. 

::: {.callout-note icon="false}

## Exercise
The code below loads the data set `cats` from the `MASS` package and makes a basic scatter plot of body weight against heart weight.

```{r plotex, echo = TRUE}
library(MASS)
plot(cats$Bwt, cats$Hwt)
```

Download and install the package `MASS` then extend the code above, to recreate the plot below (hint : you can specify `col` and `pch` as one number or a vector of numbers).

```{r plotexa, echo = FALSE}
library(MASS)
plot(cats$Bwt, cats$Hwt, pch = as.numeric(cats$Sex),
     xlab = "body weight (kg)", ylab = "heart weight (g)",
     col = as.numeric(cats$Sex))
legend("topleft", legend = c("Female", "Male"), pch = c(1, 2), col = c(1, 2))
```


:::

::: {.callout-note collapse="true" icon="false}
## Solution

```{r plotexab}
library(MASS)
plot(cats$Bwt, cats$Hwt, pch = as.numeric(cats$Sex),
     xlab = "body weight (kg)", ylab = "heart weight (g)",
     col = as.numeric(cats$Sex))
legend("topleft", legend = c("Female", "Male"), pch = c(1, 2), col = c(1, 2))
```

:::

